---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Show any code that is needed to:

1. Load the data (i.e. read.csv())

```{r echo=TRUE}
zipfile="activity.zip"
datafile="activity.csv"

# unziping with unz() command and defining NA string
data<-read.csv(unz(zipfile,datafile), stringsAsFactors=FALSE, na.strings="NA")
```

2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r echo=TRUE}

# keeping $date as strings in $dateSrtring
data$dateString<-data$date

# changing $date into date format
data$date<-as.Date(data$date)

# converting $interval into human friendly strings representing time - saving into $timeString. This will be used to build human friendly scales during ploting
data$timeString<-paste(sprintf("%02d",floor(data$interval/100)),sprintf("%02d",data$interval-floor(data$interval/100)*100),sep=":")

# converting $timeString into factor - required for dyplyr manipulations
data$timeString<-as.factor(data$timeString)
```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day

```{r echo=TRUE}
library(stats)

# using xtabs() for calculation, saving into variable for later usage
steps_per_day<-xtabs(steps~date, data=data)
steps_per_day
```

2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

```{r echo=TRUE}
# plotting histogram
hist(steps_per_day)
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r results="asis"}
mean_steps_per_day<-mean(steps_per_day)
median_steps_per_day<-median(steps_per_day)

# changin scientific notation parameters to assure proper numbers printing
options(scipen = 1, digits = 2)

```
mean of the total number of steps taken per day is `r mean_steps_per_day`

median of the total number of steps taken per day is `r median_steps_per_day`


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r echo=TRUE}
library(dplyr)

# preparing data
time_serie<-data %>% 
    select(steps,timeString) %>% # selecting columns
    group_by(timeString) %>% # grouping
    summarize(average=mean(steps, na.rm=TRUE)) #summarizing

library(lattice)

# plotting
xyplot(average ~ timeString, time_serie, type="l", scales=list(x=list(at=seq(1,12*24,24))))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}

# it can be calculated using following command
time_serie[which.max(time_serie$average),]$timeString

```

on average across all the days interval `r time_serie[which.max(time_serie$average),]$timeString`, contains the maximum number of steps


## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r echo=TRUE}
total_number_NA<-sum(is.na(data))
```

total number of missing values in the dataset is `r total_number_NA`

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Lets check where NA's are located.

1st let's separate all NAs and count them
```{r echo=TRUE}
NAs<-data[is.na(data$steps),]
nrow(NAs)
```

then let's check on which dates NA's are located and how many there are unique dates with NAs 
```{r echo=TRUE}
unique(NAs$date)
length(unique(NAs$date))

```

then let's check on which intervals NA's are located and how many there are unique intervals with NAs 

```{r echo=TRUE}
unique(NAs$interval)
length(unique(NAs$interval))

```

Number of unique dates with NAs is `r length(unique(NAs$date))`
Number of unique intervals with NAs is `r length(unique(NAs$interval))`.

It shows us that NAs are located ONLY on `r length(unique(NAs$date))` dates (on all intervals):
`r unique(NAs$date)` as:

Number of unique dates with NAs (`r length(unique(NAs$date))`) * Number of unique intervals with NAs (`r length(unique(NAs$interval))`) = `r total_number_NA`.

So NAs are only on `r length(unique(NAs$date))` dates. So my strategy proposal for filling in all of the missing values (which means all the missing dates) in the dataset is to use all daily means grouped by every 5-minute interval (it's already calculated in time_serie variable). This approach will make low impact on all data.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r echo=TRUE}

# creating new dataset
dataNONa<-data

# filling missing data
dataNONa[is.na(dataNONa$steps),]$steps<-time_serie$average
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r echo=TRUE}

# preparing data
steps_per_day_NONa<-xtabs(steps~date, data=dataNONa)
steps_per_day_NONa
# buildign histogram
hist(steps_per_day_NONa)

#calculating mean and median
mean_steps_per_day_NONa<-mean(steps_per_day_NONa)
median_steps_per_day_NONa<-median(steps_per_day_NONa)

```

mean of the total number of steps taken per day is `r mean_steps_per_day_NONa` (in the 1st part of assignment it was: `r mean_steps_per_day`)

median of the total number of steps taken per day is `r median_steps_per_day_NONa` (in the 1st part of assignment it was: `r median_steps_per_day`)

So we can see that impact is very low


## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

#i prefer not to use weekdays() function as it shows weekdays names in local language. So using this as solution on computer with different Locale could generate errors. Instead I'm using isWeekday() from "timeDate" package 
```{r echo=TRUE}
library(timeDate)

# create new variable
dataNONa$wDay<-factor(isWeekday(data$date), labels=c("weekend","weekday"))

```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r echo=TRUE}

# preparing data
time_serie_wD<-dataNONa %>% 
    select(steps,timeString,wDay) %>% # selecting colums
    group_by(wDay, timeString) %>% # grouping
    summarize(average=mean(steps, na.rm=TRUE)) # summarizing

# plotting
xyplot(average ~ timeString|wDay, time_serie_wD, type="l", scales=list(x=list(at=seq(1,12*24,24))), layout=c(1,2))

```
